---
title: "Welcome"
editor_options:
  markdown:
    wrap: 72
    
execute:
  echo: false
  message: false
  warning: false
---

This site provides access to the R package validation platform from statisticians of the
SCTO Statistics & Methodology Platform.

Statisticians of the SCTO take a risk based approach to package utilisation, whereby the risk of the product (e.g. report) defines how validated a package must be to be used in the product.

```{r}
library(gt)
library(dplyr)
data.frame(prod = "Product\nrisk", 
           prod_risk = c("Low", "Medium", "High"),
           pkg_low = "Allowed",
           pkg_medium = c("Allowed", "Allowed", "Needs assessment"),
           pkg_high = c("Allowed", "Needs assessment", "Needs assessment")) |> 
  gt(groupname_col = "prod", id = "mygt") |> 
  tab_spanner(columns  = vars(pkg_low, pkg_medium, pkg_high), 
              label = "R – package risk (baseline risk)") |> 
  cols_label(pkg_low = "Low", 
             pkg_medium = "Medium", 
             pkg_high = "High",
             prod_risk = "") |> 
  cols_width(
    2:4 ~ px(150),
  ) |> 
  tab_options(
    row_group.as_column = TRUE,
    quarto.disable_processing = TRUE
  ) |> 
  cols_align(
    align = c("center"),
    columns = everything()
  ) |> 
  data_color(columns = pkg_high, 
             method = "factor",
             palette = c("#91cf60", "#fc8d59"),
             domain = c("Allowed", "Needs assessment"),
             ) |> 
  data_color(columns = pkg_medium, 
             method = "factor",
             palette = c("#91cf60", "#fc8d59"),
             domain = c("Allowed", "Needs assessment"),
             ) |> 
  data_color(columns = pkg_low, 
             method = "factor",
             palette = c("#91cf60", "#fc8d59"),
             domain = c("Allowed", "Needs assessment"),
             ) |> 
  fmt_markdown(columns = TRUE) |> 
  tab_style(
    style = "vertical-align:middle; horizontal-align:center",
    locations = cells_row_groups()
  ) 
```





## Validated packages

```{r}
#| output: false
install.packages('validation', 
                 repos = c('https://ctu-bern.r-universe.dev', 
                           'https://cloud.r-project.org'), 
                 quiet = TRUE, verbose = FALSE)
library(validation)
library(gtExtras)
pkgs <- load_pkg_table()
```

`r nrow(pkgs)` packages or package versions have been validated by the SCTO Statistics & Methodology Platform so far.

```{r}
tmp <- pkgs |> 
  rowwise() |> 
  mutate(url = gt_hyperlink(text = "View on GitHub", url = issue_url),
         url = purrr::map(url, html),
         ) |> 
  select(package, version, final_score_cat, author, maintainer, purpose, url) # View()

risks <- unique(tmp$final_score_cat)
full_palette <- c(Low = "#91cf60", Medium = "#ffffbf", High = "#fc8d59")
observed_palette <- full_palette[risks]
tmp |> 
  gt() |> 
  cols_label(package = "Package",
             version = "Version",
             final_score_cat = "Package Risk",
             author = "Author",
             maintainer = "Maintainer",
             purpose = "Purpose",
             url = "Validation link") |> 
  opt_interactive(use_filters = TRUE,
                  use_sorting = FALSE) |>
  # harmonize table font to quarto default
  opt_table_font(font = "Source Sans Pro") |> 
  # data_color(final_score_cat, palette = c("#91cf60", "#ffffbf", "#fc8d59")) |> 
  gt_badge(final_score_cat, 
           palette = observed_palette, alpha = 1)
```

The following packages have newer versions available, which may be an indication that the validation is outdated:

```{r}
# Check for updates of installed packages
pkgs <- load_pkg_table()
# installed_packages <- installed.packages() |> as.data.frame()
available_packages <- available.packages(contriburl = "https://cran.rstudio.com/") |> 
  as.data.frame()
pkgs |> 
  dplyr::summarize(version = max(version), .by = package) |> 
  dplyr::left_join(available_packages, by = dplyr::join_by("package" == "Package")) |> #View()
  dplyr::filter(version < Version) |> 
  dplyr::select(package, version, Version) |> 
  dplyr::rename(Package = package,
         `Validated version` = version,
         `Available version` = Version
         )
```

